---
title: "`r config$publication_name`"
pagetitle: NHS Dental Statistics for England
always_allow_html: true
output:
  word_document:
    toc: yes
    toc_depth: '3'
  html_document:
    anchor_sections: no
    css: 
      - www/style.css
      - www/nhs-min.css
    toc: yes
    toc_depth: 3
    toc_float:
      collapsed: no
---

<html lang="en">

```{r setup, include=FALSE}

# set code chunk options to disable echo by default
knitr::opts_chunk$set(echo = FALSE,
                      warning = FALSE,
                      message = FALSE)

```

<div class = "quartoheader", id = "fixed-header">
  <header class="nhsuk-header" role="banner">
    <div class="nhsuk-width-container nhsuk-header__container">
      <div class="nhsuk-header__logo nhsuk-header__logo--only">
        <a class="nhsuk-header__link" href="https://www.nhsbsa.nhs.uk/" aria-label="NHSBSA home" target="_blank">
          <img class="nhsuk-logo" src="www/logo-nhsbsa.svg"" name="NHSBSA logo" alt="NHS Business Services Authority">
        </a>
      </div>
    </div>
</div>

<main>

# `r config$publication_sub_name` {.toc-ignore}

`r paste0("Published ", config$publication_date)`

::::{.row style="display: flex; padding-bottom: 10px;"} 
:::{.col-md-12} 
`r infoBox_border(header = "Feedback", text = paste0("We are interested in any feedback about the publication, which you can send by using our ", htmltools::HTML(paste("<a href='",config$stats_survey_link,"'>Official Statistics feedback survey</a>")),"."), width = "100%")` 
::: 
::::

## Key findings

::::{.row style="display: flex; padding-bottom: 10px;"} 
:::{.col-md-4} 
`r infoBox_no_border(header = "", text = "<b></b>", width = "100%")` 
::: 
:::{.col-md-4} 
`r infoBox_no_border(header = "", text = "<b></b>", width = "100%")` 
:::
:::{.col-md-4} 
`r infoBox_no_border(header = "", text = "<b></b>", width = "100%")` 
:::
::::

---

## 1.Things you should know

### 1.1. Background {.toc-ignore}

### 1.2. What is in these statistics {.toc-ignore}

### 1.3. Definitions {.toc-ignore}

::::{.row style="display: flex; margin-bottom: 24px;"} 
:::{.col-md-12 .toc-ignore}
`r infoBox_border(header = "", text = "", width = "100%")` 
::: 
::::

---

## 2. Results and commentary

---

## 3. Background

---

## 4. About these statistics

Further information on the methodology used in this publication and further background information is available in our [Background Information and Methodology]() supporting document.

---

## 5. Rounding

The high-level figures in this statistical summary have been rounded as per the table below:

```{r rounding_table}
rounding_table <- data.table(
  "From" = c("0", "1,001", "10,001", "100,001", "1,000,001", "10,000,001", "100,000,001", "10,000,000,001"),
  "To" = c("1,000", "10,000", "100,000", "1,000,000", "10,000,000", "100,000,000", "10,000,000,000", "100,000,000,000"),
  "Round to nearest" = c("1", "100", "1,000", "10,000", "100,000", "1,000,000", "10,000,000", "100,000,000")
)

knitr::kable(rounding_table, align = "rrr")
```

All changes and totals are calculated prior to rounding. Percentage changes are calculated prior to rounding and then are rounded to the nearest whole number. As all figures within this statistical summary have been rounded, they may not match totals elsewhere when aggregated.

The summary tables released with this publication allow users to investigate this data at lower levels of granularity. Figures in the supplementary tables have not been rounded.

---

## 6. Statistical disclosure control

Statistical disclosure control has been applied to these statistics. Further information about our statistical disclosure control protocol [can be found on our website](https://www.nhsbsa.nhs.uk/policies-and-procedures).

The high-level figures in this statistical summary have been rounded where appropriate for clarity. This is to make this narrative as accessible as possible to all readers. The summary tables released with this publication allow users to investigate this data at lower levels of granularity. Figures in the supplementary tables have not been rounded.

---

## 7. Accessibility

If you need information on this website in a different format, such as accessible PDF, large print, easy read, audio recording or braille, you can contact us by:

**Email**: accessibility@nhsbsa.nhs.uk

**Telephone**: 0191 203 5318

[Find out about call charges](https://www.nhsbsa.nhs.uk/contact-us/call-charges-and-phone-numbers)

We will consider your request and get back to you in 5 working days.

These contact details are only for accessibility queries. This email address is not for technical queries or IT problems. If you have a query that is not about accessibility, go to the ‘Feedback and contact us’ section of this page.

Read our [Accessibility statement for Official Statistics Narratives](https://www.nhsbsa.nhs.uk/accessibility-statement-official-statistics-narratives).

---

## 8. Feedback and contact us

Feedback is important to us. We welcome any questions and comments relating to these statistics. 

You can complete a [short survey about this publication](`r config$stats_survey_link`) to help us improve the Official Statistics that we produce. All responses will remain anonymous, and individuals will not be identifiable in any report that we produce.

You can view our [privacy policy](https://www.nhsbsa.nhs.uk/our-policies/privacy) on our website to see how your data is used and stored.

You can contact us by:

**Email:** statistics@nhsbsa.nhs.uk

**You can also write to us at:**

NHSBSA - Statistics  
NHS Business Services Authority  
Stella House  
Goldcrest Way  
Newburn Riverside  
Newcastle upon Tyne  
NE15 8NY

**Responsible statistician:** `r config$responsible_statistician`

</main>
